Old or Heavy?\\Decaying Gracefully with Age/Weight Shapes.

### 2019-03-01 papers/CADE19/paper.tex
updates

\documentclass{llncs}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage[dvipsnames]{xcolor}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{todonotes}
%\newcommand{\todo}[1]{}

@@

Vampire is a portfolio solver and is typically run in a mode that attempts multiple different \emph{strategies} in quick succession e.g. in a 30 second run it may attempt 10 or more different strategies. These strategies employ many different options including different saturation algorithms (including Otter and Discount), preprocessing options, literal selection strategies, inference rules, and clause selection heuristics. The portfolio mode is a significant improvement on any single strategy.

@@

%%%%%%
\paragraph{Related Work.} We briefly review the clause selection approaches taken by other solvers. 
%
Otter \cite{} and Prover9 \cite{} CHECK BUT THEY PROBABLY  ALLOW BOTH.
%
E~\cite{E} allows the user to specify an arbitrary number of priority queues and a weighted round-robin scheme that determines how many clauses are picked from each queue. The default is to use a combination of age and weight selection although there is also a complex strategy developed by a genetic algorithm~\cite{DBLP:conf/gcai/SchaferS15}. 
%
SPASS~\cite{Spass} uses symbol-counting based clause selection. 
%
iProver~\cite{} follows in E in having a number of configurable queues but relies mainly on age and weight heuristics in those queues. 
%
The general idea in this paper of \emph{vary the ratio over time} can be applied to any ratio-based clause selection strategy and our specific results apply to those that take a ratio between age and weight.

@@

The second result is that there is no ``best'' AWR across this full set of problems. Table \ref{tab:no-best-awr} shows, for each AWR value, the \% of problems solved, the number solved uniquely, the maximum and mean drop in performance. The drop in performance is the how many times more activations were required for a proof in the worst/average case compared to the best. No AWR value solves all problems. 
A ratio of 1:4 produces an unusually well-behaved maximum performance drop.
\sandm~found that 1:5 had a similar property, but this might be explained by differences in prover and test environment. It is interesting to note that the extreme AWR values solve fewer problems overall but solve the most uniquely. This is typically in saturation-based proof search; approaches that do not perform well in general may perform well in specific cases where the general approach does not.

@@

Although choosing a good AWR value is important, this is covered in part by the use of strategy scheduling~\cite{CADE18} in which many AWR values are tried in sequence (along with other prover options).
Additionally, given that varying the AWR can have such a large impact, it seems likely that a constant AWR fixed for the entire proof search is unlikely to be optimal for any given problem.
This can be shown by running Vampire with a randomised sequence of age/weight ratios given by a random walk repeatedly, then finding the best after a large number of repetitions.
Applying this method with 10,000 repetitions to the problem seen earlier (\texttt{PRO017+2}) yields the example AWR trend shown in Figure \ref{fig:random-walk}, which reduces the best number of activations from 330 with a fixed AWR, to 287 with a varying AWR.

Unsurprisingly, in ad-hoc experiments on other problems, the best shape is rarely constant.
This suggests that implementing other shapes, such as an increasing or decreasing trend, might lead to quicker proofs in the Vampire theorem prover.

@@

%%%%%%%
\subsection{Varying AWR in Vampire} \label{sec:varying:implementation}

If changing AWR is sometimes useful for finding proofs more quickly, this could be useful as a Vampire option.
In general we would like to describe any possible sequence that the AWR could follow during proof search.
However, some details constrain the design space:
\begin{enumerate}
	\item Changing the AWR too frequently or sharply has little effect, due to the ``balancing'' algorithm --- see Section \ref{sec:introduction}.
	\item A general (configurable) \emph{shape} is more likely to be widely applicable than a specific series of data points.
	\item The shape must extend naturally to an indefinitely-long proof search.
\end{enumerate}

In this work we selected two general trends to explore: a trend away from the original (fixed) AWR toward 1:1 (``decay''), and a trend from 1:1 toward the original setting (``converge'').
Since a simple linear shape does not extend well to indefinite proof search (it is unclear what should happen after either 1:1 or the target AWR is reached), an exponential decay function is used instead.
These exponential shapes are further parameterised by an integral \emph{shape frequency} setting, which controls the rate of decay or convergence: every \(n\) steps, the difference between the current and the target AWR is halved, rounding where necessary.
In future, this might allow the use of repeating patterns such as a sinusoid, hence \emph{frequency}.
Figure \ref{fig:decay-and-converge} shows rates at which the new configurations converge or decay from the fixed AWR setting for some indicative frequency settings.

@@

Together, these new settings allow a number of new option combinations, which can be used in conjunction with Vampire's portfolio mode pending integration into the strategy schedules.
This version of the prover is currently in a separate branch in the main Vampire source repository\footnote{\texttt{https://github.com/vprover/vampire/tree/awr-shapes}}.

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experimental Evaluation}
\label{sec:experiments}
\todo{Only mention CASC once and refer to portfolio mode elsewhere.}
An experimental evaluation of this approach on the TPTP\footnote{version 7.1.0} benchmark~\cite{tptp} follows.
Vampire first ran to establish baseline performance in its competition \emph{CASC mode} on all problems in TPTP, with a wallclock time limit of 300 seconds.
For experimental purposes, the same environment was preserved with the exception of activating the new ``shapes'' settings with a range of different shape frequencies.

\subsection{Benchmark Performance}
The baseline solved 13,057 problems in total.
No experimental configuration improved on this figure, but some problems not solved by baseline were solved by the new configurations, and some entirely new problems were solved.
The union of all \emph{converge} and \emph{decay} configurations improved on the baseline, with 13,167 and 13,106 solved problems respectively.

@@

\begin{table}
	\caption{
Results for the tested configurations.
\emph{Proved} refers to the total number of problems a configuration solved.
\emph{Fresh} is the number of problems a configuration solved which were not solved by the baseline.
\emph{Uniques} is the number of problems a configuration solved which were not solved by any other configuration.
\emph{\(u\)-score} is a refined unique score which correlates to a configuration's utility in solving new problems, as used in Hoder \emph{et al.}~\cite{u-score}.
	}
	\centering
	\begin{tabular}{l l l l l l}
		Configuration & Frequency & Proved & Fresh & Uniques & \(u\)-score\\
		\hline
		baseline & -- & 13057 & 0 & 1 & 714.2\\
		converge & 1 & 13039 & 24 & 3 & 714.3\\
		converge & 5 & 13029 & 27 & 1 & 709.5\\
		converge & 10 & 13028 & 35 & 5 & 714.3\\
		converge & 50 & 13015 & 45 & 5 & 712.8\\
		converge & 100 & 12976 & 51 & 1 & 705.9\\
		converge & 500 & 12895 & 63 & 4 & 698.3\\
		converge & 1000 & 12837 & 52 & 0 & 688.6\\
		converge & 5000 & 12775 & 53 & 1 & 682.4\\
		converge & 10000 & 12751 & 53 & 0 & 678.7\\
		decay & 1 & 12698 & 48 & 1 & 673.6\\
		decay & 5 & 12702 & 51 & 1 & 674.9\\
		decay & 10 & 12698 & 48 & 1 & 674.2\\
		decay & 50 & 12712 & 49 & 2 & 679.1\\
		decay & 100 & 12726 & 46 & 1 & 678.8\\
		decay & 500 & 12795 & 29 & 1 & 685.5\\
		decay & 1000 & 12860 & 29 & 2 & 692.6\\
		decay & 5000 & 12982 & 16 & 2 & 707.1\\
		decay & 10000 & 13002 & 7 & 0 & 706.3\\
		\hline
		converge & (combined) & 13167 & 117 & 41 & --\\
		decay & (combined) & 13106 & 93 & 17 & --\\
	\end{tabular}
	\label{tab:results}
\end{table}

\subsection{New Problems}
Some problems were solved which were marked with an ``Unknown'' status in the TPTP headers.
Vampire baseline (and several configurations) were able to solve \texttt{GEO120+1}, but only converging with frequency 50 solved \texttt{SET345-6} and only decaying with frequency 1 solved \texttt{LAT320+3}.

\clearpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions and Future Work}
\label{sec:conclusions}
Age/weight ratios have a significant effect on the performance of proof search in the Vampire theorem prover, but there is no clear optimal age/weight ratio.
Further, we have demonstrated that varying the age/weight ratio over time \emph{during} proof search can improve further on an optimal, but fixed age/weight ratio in terms of the number of activations.
Experiments within Vampire on the TPTP benchmark suggest that these \emph{age/weight shapes} show promise for future developments in this novel approach to proof search.

Future directions for research in this area include trying a greater number of ``shapes'' (such as repeating patterns), other approaches for parameterising these shapes, a pseudo-random approach to age/weight instead of the balancing algorithm, and integration of the new approaches into existing strategy schedules.
\bibliographystyle{plain}
\bibliography{references}
\end{document}

>>>>

\documentclass{llncs}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage[dvipsnames]{xcolor}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{todonotes}
%\usepackage{multicol}
%\newcommand{\todo}[1]{}

@@

Vampire is a portfolio solver and is typically run in a mode that attempts multiple different \emph{strategies} in quick succession, e.g. in a 30 second run it may attempt 10 or more different strategies, and may run these in parallel with different priorities~\cite{DBLP:conf/cade/RawsonR18}. These strategies employ many different options including different saturation algorithms (including Otter and Discount), preprocessing options, literal selection strategies, inference rules, and clause selection heuristics. The portfolio mode is a significant improvement on any single strategy.

@@

%%%%%%
\paragraph{Related Work.} We briefly review the clause selection approaches taken by other solvers. 
%
Otter \cite{} and Prover9 \cite{} CHECK BUT THEY PROBABLY  ALLOW BOTH.
%
E~\cite{E} allows the user to specify an arbitrary number of priority queues and a weighted round-robin scheme that determines how many clauses are picked from each queue. The default is to use a combination of age and weight selection although there is also a complex strategy developed by a genetic algorithm~\cite{DBLP:conf/gcai/SchaferS15}. 
%
SPASS~\cite{Spass} uses symbol-counting based clause selection. 
%
iProver~\cite{Iprover} follows in E in having a number of configurable queues but relies mainly on age and weight heuristics in those queues. 
%
The general idea in this paper of \emph{vary the ratio over time} can be applied to any ratio-based clause selection strategy and our specific results apply to those that take a ratio between age and weight.

@@

The second result is that there is no ``best'' AWR across this full set of problems. Table \ref{tab:no-best-awr} shows, for each AWR value, the \% of problems solved, the number solved uniquely, the maximum and mean drop in performance. The drop in performance is the how many times more activations were required for a proof in the worst/average case compared to the best. No AWR value solves all problems, with the best being 1:5. 
A ratio of 1:4 produces an unusually well-behaved maximum performance drop.
\sandm~found that 1:5 had a similar property, but this might be explained by differences in prover and test environment. It is interesting to note that the extreme AWR values solve fewer problems overall but solve the most uniquely. This is typically in saturation-based proof search; approaches that do not perform well in general may perform well in specific cases where the general approach does not.

@@

Although choosing a good AWR value is important, this is covered in part by the use of strategy scheduling in which many AWR values are tried in sequence (along with other prover options).
Additionally, given that varying the AWR can have such a large impact, it seems likely that a constant AWR fixed for the entire proof search is unlikely to be optimal for any given problem.
This can be shown by running Vampire with a randomised sequence of age/weight ratios given by a random walk repeatedly, then finding the best after a large number of repetitions.
Applying this method with 10,000 repetitions to the problem seen earlier (\texttt{PRO017+2}) yields the example AWR trend shown in Figure \ref{fig:random-walk}, which reduces the best number of activations from 330 with a fixed AWR, to 287 with a varying AWR. 
%
Unsurprisingly, in ad-hoc experiments on other problems, the best shape is rarely constant.
This suggests that implementing other shapes, such as an increasing or decreasing trend, might lead to quicker proofs in the Vampire theorem prover.

@@

%%%%%%%
\subsection{Varying AWR (in Vampire)} \label{sec:varying:implementation}

We describe how we vary AWR dynamically and how this is achieved in Vampire. 
%
%If dynamically varying AWR is sometimes useful for finding proofs more quickly, this could be useful as a Vampire option.
In general we would like to describe any possible sequence that the AWR could follow during proof search.
However, some details constrain the design space:
\begin{enumerate}
	\item Changing the AWR too frequently or sharply has little effect, due to the ``balancing'' algorithm --- see Section \ref{sec:introduction}.
	\item A general (configurable) \emph{shape} is more likely to be widely applicable than a specific series of data points.
	\item The shape must extend naturally to an indefinitely-long proof search.
\end{enumerate}

In this work we select two general trends to explore: a trend away from the original (fixed) AWR toward 1:1 (``decay''), and a trend from 1:1 toward the original setting (``converge''). In our investigations we found that even fluctuating sequences had a general trend and these two fixed trends are reasonable approximations of this. We have chosen to use an `original' AWR value to move away from/towards as this original AWR has already been shown to be `useful' in some sense.

Since a simple linear shape does not extend well to indefinite proof search (it is unclear what should happen after either 1:1 or the target AWR is reached), an exponential decay function is used instead.
These exponential shapes are further parameterised by an integral \emph{shape frequency} setting, which controls the rate of decay or convergence: every \(n\) steps, the difference between the current and the target AWR is halved, rounding where necessary.
In future, this might allow the use of repeating patterns such as a sinusoid, hence \emph{frequency}.
Figure \ref{fig:decay-and-converge} illustrates rates at which the new configurations converge or decay from the fixed AWR setting for some indicative frequency settings.

@@

We introduce two new options: \texttt{--age\_weight\_ratio\_shape} can take the values \emph{constant}, \emph{decay}, or \emph{converge} and selects one of the above shapes; and \texttt{--age\_weight\_ratio\_shape\_frequency} specifies the frequency (rate) or convergance/decay (default is 100). These are used with the existing \texttt{--age\_weight\_ratio} option (default 1:1) to give a number of new option combinations, which can be used in conjunction with Vampire's portfolio mode pending integration into the strategy schedules.
This version of the prover is currently in a separate branch in the main Vampire source repository\footnote{\texttt{https://github.com/vprover/vampire/tree/awr-shapes}}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experimental Evaluation}
\label{sec:experiments}

We perform two experiments to evaluate the new techniques. The first compares the various options attempting to draw some conclusions about which option values work well together. The second looks at how useful the new options are in the context of portfolio solving. Both experiments use the TPTP (version 7.1.0) benchmark~\cite{tptp} and were run on StarExec \cite{starexec}.

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparing New Options}

\begin{table}[t]
\caption{Number of problems solved (top) and unique problems solved (bottom) by various configurations varying AWR, AWR shape, and AWR frequency.\label{tab:short:runs}}
\centering
\begin{tabular}{l|l|lllll|lllll}
&&  \multicolumn{5}{|c}{converge} &  \multicolumn{5}{|c}{decay} \\  
AWR & constant & \multicolumn{5}{|c|}{Frequency}  & \multicolumn{5}{|c}{Frequency} \\  
&& 1 & 10 & 100 & 1000 & Union & 1 & 10 & 100 & 1000 &Union  \\ \hline
\multicolumn{10}{c}{Problems Solved} \\ \hline
10:1		&5911 & 5908&5910 & 5969 &6135 &  & 5908 &5910  & 5969  &6135  \\
1:10		&6417& 6412&6421 &6389 & 6325  \\
1:100	&6026& 6028& 6010& 5971& 6148\\
1:1000	&5387&5395  &5408 &5510 &6019  \\ \hline
Union & \\ \hline \hline
\multicolumn{10}{c}{Uniquely Solved} \\ \hline
10:1		&1 & 4&2 & 5 &10 & 4&2 & 5&10 \\
1:10		& \\
1:100	& \\
1:1000	&\\ \hline
Union & \\ 
\end{tabular}
\end{table}

We ran Vampire in default mode (with {\tt discount} saturation algorithm) for 10s whilst varying  \texttt{age\_weight\_ratio} and \texttt{age\_weight\_ratio\_shape\_frequency} for both new values of \texttt{age\_weight\_ratio\_shape}. We also ran the constant default AWR of 1:1 as a baseline.

The results are given in Table~\ref{tab:short:runs}. The results for the different shapes are grouped into columns and then by frequency with rows giving results per AWR value. We separately report the total number of problems solved and those solved uniquely (by a single strategy across all experiments).

The best combination of options overall was X. In general, strategies that converged performed better/worse than those that diverged.

In total X out of all Y strategies contributed some unique solutions. This shows that these new options are complementary in general.

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Contribution to Portfolio}

\begin{table}[t]
	\caption{
Results for the tested configurations.
\emph{Proved} refers to the total number of problems a configuration solved.
\emph{Fresh} is the number of problems a configuration solved which were not solved by the baseline.
\emph{Uniques} is the number of problems a configuration solved which were not solved by any other configuration.
\emph{\(u\)-score} is a refined unique score which correlates to a configuration's utility in solving new problems, as used in Hoder \emph{et al.}~\cite{u-score}.
	}
	\centering
	\begin{tabular}{l l l l l l}
		Configuration & Frequency & Proved & Fresh & Uniques & \(u\)-score\\
		\hline
		baseline & -- & \textbf{13,057} & 0 & 1 & 714.2\\ \hline
		converge & 1 & 13,039 & 24 & 3 & 714.3\\
		converge & 5 & 13,029 & 27 & 1 & 709.5\\
		converge & 10 & 13,028 & 35 & 5 & 714.3\\
		converge & 50 & 13,015 & 45 & 5 & 712.8\\
		converge & 100 & 12,976 & 51 & 1 & 705.9\\
		converge & 500 & 12,895 & 63 & 4 & 698.3\\
		converge & 1000 & 12,837 & 52 & 0 & 688.6\\
		converge & 5000 & 12,775 & 53 & 1 & 682.4\\
		converge & 10000 & 12,751 & 53 & 0 & 678.7\\ \hline
		decay & 1 & 12,698 & 48 & 1 & 673.6\\
		decay & 5 & 12,702 & 51 & 1 & 674.9\\
		decay & 10 & 12,698 & 48 & 1 & 674.2\\
		decay & 50 & 12,712 & 49 & 2 & 679.1\\
		decay & 100 & 12,726 & 46 & 1 & 678.8\\
		decay & 500 & 12,795 & 29 & 1 & 685.5\\
		decay & 1000 & 12,860 & 29 & 2 & 692.6\\
		decay & 5000 & 12,982 & 16 & 2 & 707.1\\
		decay & 10000 & 13,002 & 7 & 0 & 706.3\\
		\hline
		converge & (combined) & 13,167 & 117 & 41 & --\\
		decay & (combined) & 13,106 & 93 & 17 & --\\
	\end{tabular}
	\label{tab:results}
\end{table}

Our next experiment aims to answer the question ``how much can the portfolio mode of Vampire improve using these new options''. To address this question we use the portfolio mode used in the most recent CASC competition~\cite{CADE18} as a baseline and run the new options on top of this. Note that this portfolio mode contains techniques completely unrelated to the age/weight ratio e.g. finite model building~\cite{DBLP:conf/sat/Reger0V16}.

Vampire first ran to establish baseline performance in the given portfolio mode on all problems in TPTP, with a wallclock time limit of 300 seconds. We then forced the new options on top of the portfolio mode, using the existing AWR values in the various strategies as the starting point. The purpose is to gauge what impact adding such options to a new portfolio mode could have. In this experiment we are mainly hoping to find new solved problems and identify new strategies that could be added to a portfolio mode. Therefore, it makes sense to consider the union of all experiments ran.
%For experimental purposes, the same environment was preserved with the exception of activating the new ``shapes'' settings with a range of different shape frequencies.

Overall, the baseline solved the most problems (13,057)
No experimental configuration improved on this figure, but some problems not solved by baseline were solved by the new configurations, and some entirely new problems were solved.
The union of all \emph{converge} and \emph{decay} configurations improved on the baseline, with 13,167 and 13,106 solved problems respectively.

@@

%\subsection{New Problems}
Two problems were solved which were marked with an ``Unknown'' status in the TPTP headers.
Only converging with frequency 50 solved \texttt{SET345-6} and only decaying with frequency 1 solved \texttt{LAT320+3}.

%\clearpage

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparison with other Solvers}

\begin{table}[t]
\caption{Total number of problems solved compared to other solvers.\label{tab:other}}
\centering
\begin{tabular}{l|l|l}
Solver 		& Total solved & Uniquely solved \\ \hline
Vampire (old)	& 13,057 & \\
Vampire (new)   & 13,191 & \\
E 			& \\
iProver		& \\
CVC4		& \\
\end{tabular}
\end{table}

To place these results in context, we compare the overall number of problems solved by our new strategies with the results of other solvers (using their previous CASC competition strategy). These results are from 300s and are given in Table~\ref{tab:other}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions and Future Work}
\label{sec:conclusions}

Clause selection is a key part of any saturation-based theorem prover and age/weight ratios have a significant effect on the performance of proof search in the Vampire theorem prover. We have supported the known folklore that there is no clear optimal age/weight ratio.
Further, we have demonstrated that varying the age/weight ratio over time \emph{during} proof search can improve further on an optimal, but fixed age/weight ratio in terms of the number of activations.
Experiments within Vampire on the TPTP benchmark suggest that these \emph{age/weight shapes} show promise for future developments in this novel approach to proof search. Indeed, including our relatively simple shapes already leads to significant performance gains.

Future directions for research include trying a greater number of ``shapes'' (such as repeating patterns), other approaches for parameterising these shapes, a pseudo-random approach to age/weight instead of the balancing algorithm, and integration of the new approaches into existing strategy schedules.
\bibliographystyle{plain}
\bibliography{references}
\end{document}
